Challenges in Developing Great Quasi-Monte Carlo Software

### 2023-07-12 MCQMC2022ChallengesQMCSoftwareArticle_revision_1.tex
Fred 2023 Mac 14 Updates

@@

Lattice and digital sequences are not unique, and improved or alternate versions continue to be constructed.  Stephen Joe and Frances Kuo \cite{JoeKuo03,KuoJoe08a,SobolDirection} have proposed  Sobol' generator direction numbers, which are now widely used, but were not part of early LD sequence software.  Pierre L'Ecuyer and collaborators have created LatNet Builder \cite{LEcEtal22a} to construct good lattice and polynomial lattice sequences based on user-defined criteria. During lunch at MCQMC 2022, it was proposed that we  agree on a consistent format for storing the parameters that define good LD sequences so that they can be shared across software libraries. \SCComment{Changed LatNet Builder's reference in response to Referee's \#3, but the citation relies on another reference ArnGilKel22a which is not found}

@@

\subsection{Integrands and Variable Transformations} \label{CDHJS:sec:integrands}
The original integral or expectation arising from an application is defined in terms of the integrand, $g$, and the non-negative weight, $\lambda$.  For example,
\begin{itemize}
	\item $g$ is the discounted payoff of a financial derivative and $\lambda$ is the probability density function (PDF) for a discretized Brownian motion \cite{Gla03},

@@

To put $\mu$ into the form that is directly accessible to QMC methods requires a variable transformation, $\bsPsi$, such that
\begin{equation*}
    \bsPsi((0,1)^d) = \calT, \qquad f(\bsx) = g(\bsPsi(\bsx)) \lambda(\bsPsi(\bsx)) \abs{\frac{\partial \bsPsi}{\partial\bsx}}.
\end{equation*}
Such a transformation is typically non-unique, and the choice of a good one is equivalent to importance sampling.  If $\lambda$ is a PDF for a random variable $\bsT$, then $\mu = \EE[g(\bsT)]$.  In this case, one might choose $\bsPsi$ such that $\bsT \sim \bsPsi(\bsX)$, in which case $\lambda(\bsPsi(\bsx)) \abs{\partial \bsPsi/\partial\bsx} = 1$ and $f(\bsx) = g(\bsPsi(\bsx))$. Software such as SciPy and QMCPy facilitate such variable transformations by providing default $\bsPsi$ for common $\bsT$. This alleviates the burden on the user to determine a transformation and implement the Jacobian.

\begin{figure}[t]
    \centering
    \raisebox{0.5cm}{\includegraphics[width=0.4\textwidth]{BeamDrawing.png}}
    \includegraphics[width=0.5\textwidth]{cantileveredbeamwords.eps}
    \includegraphics[width=1\textwidth]{ldparallelbeam.eps}
    \caption{The expected deflection (upper left, image reproduced from~\cite{ParSee22a} with permission) of a cantilevered beam (upper right, \SCComment{what is $n$ for this figure?}) with random Young's modulus.  QMCPy computes the answer in serial and in parallel using UM-Bridge and docker containers.  For small error tolerances, the time and sample size is $\calO(\varepsilon^{-1})$ (lower right), and parallel processing gives a speed-up factor of four to five (lower left). These plots may be reproduced using the Jupyter notebook \cite{MCQMC2022Figs}.  }
    \label{CDHJSfigbeam}
\end{figure}

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \LSComment{\cite{UMBridge23}}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$   \LSComment{, and their derivatives if applicable. UM-Bridge allows coupling of UQ codes to models regardless of the programming languages.}

We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam \SCComment{of length~$L$}, with a random Young's modulus \SCComment{$m(x)$, assumed to be piecewise constant over three regions, each of which with a uniform distribution of uncertainty~\cite{ParSee22a}, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  The beam position $x$ is discretized at $N=31$ equidistant points, and
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. }
%approximates $u(x)$ and $m(x)$ at the nodes, respectively.
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$.
\LSComment{The beam model was written in C++ using the MUQ's model framework~\cite{MUQ2}. It is one of the UM-Bridge's benchmark problem library, available as a container.}
QMCPy simulates the random Young's modulus and computes the expected deflection of the beam at the node points for a sequence of error tolerances using lattice rules.  By taking advantage of \LSComment{clusters and containers}, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation. \LSComment{As a result, thread-parallel QMCPy can offload costly model evaluations to remote clusters~\cite{seelinger2023lowering}.}

@@

In spite of the desire for efficiency, great QMC software will be written in a variety of computer languages since different communities tend to favor different languages. Trade-offs will be made between efficiency and convenience.

@@

\SCComment{In QMCPy, we recognize the crucial role of non-code contributors to the project~\cite{K23}. Besides coding, we also need documentation, data visualization,  testing, user interface or experience design, web content development, community management, and release management. These non-code contributions help make the project more  accessible. Non-coding roles offer opportunities for individuals with less coding experiences to develop technical and non-technical skills over time.  Additionally, the emergence of no-code or low-code analytical software such as RapidMiner, KNIME, Orange, SAS Studio, or SAS Viya enable individuals less experienced with complex coding to learn and contribute in a user-friendly environment.  We need and welcome non-code contributors to work alongside code contributors for QMCPy, enhancing the overall health of the project and the diversity of the community. }

@@

\SCComment{Response to comment \#2: In this paper, we have focused predominantly on employing QMC sampling in estimating the expectations of functions. However, it can be extended to other important mathematical problems such as function approximation, simulation of Markov chains,  density estimation, and many MC-based randomized algorithms. For instance, Array-RQMC has proven invaluable for maintaining and sorting arrays of Markov chains \cite{l2018sorting,puchhammer2021variance}.
The SSJ library~\cite{l2002ssj} is a comprehensive tool kit that includes a wide range of QMC point sets, sequences, and randomizations in the ``hups'' and ``mcqmctools'' packages. It will be of great interest to the QMCPy community to make connections to these and other well established software in order to become applicable to an increasingly wider scope of problems and applications of computational mathematics and statistics.
}

\SCComment{Response to comment \#4:  Python is known for its versatility, but it can be slower than highly optimized implementations in languages like Java or C++.  For QMCPy developers, it would be beneficial to investigate such underlying interfaces or consider other code acceleration techniques by leveraging high-performance libraries such as Numba, JAX, Joblib, or Ray in the future.
}

\SCComment{Question: Do we have only one referee's report for this paper?}

\section*{Acknowledgements}
\SCComment{The authors are grateful to the referee's constructive comments. We also want to thank Linus Seelinger for feedback and discussion. }

>>>>

\bigskip

@@

Lattice and digital sequences are not unique, and improved or alternate versions continue to be constructed.  Stephen Joe and Frances Kuo \cite{JoeKuo03,KuoJoe08a,SobolDirection} have proposed  Sobol' generator direction numbers, which are now widely used, but were not part of early LD sequence software.  Pierre L'Ecuyer and collaborators have created LatNet Builder \cite{LEcEtal22a} to construct good lattice and polynomial lattice sequences based on user-defined criteria. During lunch at MCQMC 2022, it was proposed that we  agree on a consistent format for storing the parameters that define good LD sequences so that they can be shared across software libraries.

@@

\subsection{Integrands and Variable Transformations} \label{CDHJS:sec:integrands}
The original integral or expectation arising from an application is defined in terms of the integrand, $g :\mathcal{T} \to \R$, and the non-negative weight, $\lambda$, as described in \eqref{eq:cubSummary}.  For example,
\begin{itemize}
	\item $g$ is the discounted payoff of a financial derivative and $\lambda$ is the probability density function (PDF) for a discretized Brownian motion \cite{Gla03},

@@

To put $\mu$ into the form that is directly accessible to QMC methods requires a variable transformation, $\bsPsi$, such that
\begin{equation*}
    \bsPsi((0,1)^d) = \calT, \qquad f(\bsx) = g(\bsPsi(\bsx)) \lambda(\bsPsi(\bsx)) \abs{\frac{\partial \bsPsi}{\partial\bsx}}.
\end{equation*}
Such a transformation is typically non-unique, and the choice of a good one is equivalent to importance sampling.  If $\lambda$ is the PDF for a random variable $\bsT$, then $\mu = \EE[g(\bsT)]$.  In this case, one might choose $\bsPsi$ such that $\bsT \sim \bsPsi(\bsX)$, $\bsX \sim \calU(0,1)^d$, in which case $\lambda(\bsPsi(\bsx)) \abs{\partial \bsPsi/\partial\bsx} = 1$ and $f(\bsx) = g(\bsPsi(\bsx))$. Software such as SciPy and QMCPy facilitate such variable transformations by providing default $\bsPsi$ for common $\bsT$. This alleviates the burden on the user to determine a transformation and implement the Jacobian.

\begin{figure}[t]
    \centering
    \raisebox{0.5cm}{\includegraphics[width=0.4\textwidth]{BeamDrawing.png}}
    \includegraphics[width=0.5\textwidth]{cantileveredbeamwords.eps}
    \includegraphics[width=1\textwidth]{ldparallelbeam.eps}
    \caption{The expected deflection (upper left, image reproduced from~\cite{ParSee22a} with permission) of a cantilevered beam (upper right) with random Young's modulus.  QMCPy computes the answer in serial and in parallel using UM-Bridge and docker containers.  For small error tolerances, the time and sample size is $\calO(\varepsilon^{-1})$ (lower right), and parallel processing gives a speed-up factor of four to five (lower left). These plots may be reproduced using the Jupyter notebook \cite{MCQMC2022Figs}.  }
    \label{CDHJSfigbeam}
\end{figure}

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \cite{UMBridge23}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$, and their derivatives if applicable. UM-Bridge allows coupling of UQ codes to models regardless of the programming languages.
{eq:cubSummary}
We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam of length~$L$, with a random Young's modulus $m(x)$, assumed to be piecewise constant over three regions, each of which with a uniform distribution of uncertainty~\cite{ParSee22a}, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  The beam position $x$ is discretized at $N=31$ equidistant points, and
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. }
%approximates $u(x)$ and $m(x)$ at the nodes, respectively.
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$.
The beam model was written in C++ using the MUQ's model framework~\cite{MUQ2}. It is one of the UM-Bridge's benchmark problem library, available as a container.

QMCPy simulates the random Young's modulus and computes the expected deflection of the beam at the node points for a sequence of error tolerances using lattice rules.  By taking advantage of clusters and containers, UM-Bridge provides automatic load balancing to facilitate seamless parallel model evaluation. As a result, thread-parallel QMCPy can offload costly model evaluations to remote clusters~\cite{seelinger2023lowering}.

@@

In spite of the desire for efficiency, great QMC software will be written in a variety of computer languages since different communities tend to favor different languages. Trade-offs will be made between efficiency and convenience.
For example, as noted by a referee, QMCPy is not particularly fast at generating tens of thousands of replicates of randomized QMC sample means in comparison to SSJ, which is based on Java. Such cases may point to the need for a Python layer over a faster implementation.

@@

We recognize the crucial role of non-code contributors to the project~\cite{K23}. Besides coding, we also need documentation, data visualization, testing, user interface or experience design, web content development, community management, and release management. These non-code contributions help make the project more  accessible. Non-coding roles offer opportunities for individuals with less coding experiences to develop technical and non-technical skills over time.  Additionally, the emergence of no-code or low-code analytical software such as RapidMiner, KNIME, Orange, SAS Studio, or SAS Viya enable individuals less experienced with complex coding to learn and contribute in a user-friendly environment.
%We need and welcome non-code contributors to work alongside code contributors for QMCPy, enhancing the overall health of the project and the diversity of the community.

@@

In this paper, we have focused predominantly on employing QMC sampling in estimating the expectations of functions. However, it can be extended to other important mathematical problems such as function approximation \cite{Kmmerer2015, PotSch21}, simulation of Markov chains \cite{l2018sorting,puchhammer2021variance},  density estimation \cite{LEcuyer2022b}.
The SSJ library~\cite{l2002ssj} is a comprehensive tool kit that includes a wide range of QMC point sets, sequences, and randomizations in the ``hups'' and ``mcqmctools'' packages. It will be of great interest to the QMCPy community to make connections to these and other well established software in order to become applicable to an increasingly wider scope of problems and applications of computational mathematics and statistics.

Python is known for its versatility, but it can be slower than highly optimized implementations in languages like Java or C++.  For QMCPy developers, it would be beneficial to investigate such underlying interfaces or consider other code acceleration techniques by leveraging high-performance libraries such as Numba, JAX, Joblib, or Ray in the future.

\section*{Acknowledgements}
The authors are grateful to the referee's constructive comments. We also want to thank Linus Seelinger for feedback and discussion. The authors acknowledge many students and collaborators who have provided insight in our journey to develop great QMC software.
